# Hierarchical GAM on fish population trends

Model we are attempting to build:

$$N_{i,t} = f(t, \alpha_i) + g(\lambda_i, \beta_i) + \varepsilon_i$$

Where:

- $N$ : population size estimate
- $i$ : population
- $t$ : time step

- $f(t, \alpha_i)$ : function estimating the temporal trend in population size
  - $\alpha$: estimated parameter describing each population's temporal slope
- $g(\lambda, \beta_i)$ : function estimating latent variation in population size
  - $\lambda$ : latent variable (to capture covariation between populations)
  - $\beta$ : estimated parameter describing each population's latent variation
- $\varepsilon$: residual variation

Because the model also estimates latent variation, we are accounting for the covariation between populations when estimating $N$. 

This model then contains key information to calculate and extend an index like the Living Planet Index, to show the average change in population size from a baseline value of 1 in the start year.

The first part of the model estimates the temporal trend of each population's size ($\alpha_i$), which contains information about population-level change _and_ community-level change:

$$\alpha_i \sim N(\mu_{\alpha}, \sigma^2_{\alpha}) $$

where:
- \mu_{\alpha} is the average population trend
- \sigma^2_{\alpha} is the variance in this average trend, giving us a measure of how variable the population trends are within the sample.

To calculate the index of change from a baseline of 1 (as in the LPI), the average population trend can then be plugged in to the LPI calculation as follows:

$$ I_t = {\mu_{\alpha}}*t + 1 $$

This trend can then be visualized and communicated with its known ($\sigma^2_{\alpha}$).


```{r, include = FALSE}
# Script to build hierarchical GAM on fish population trends

library(here)
library(dplyr)
library(tidyr)
library(mgcv)
library(tidymv)
library(gratia)
library(ggplot2)
library(ggpubr)

knitr::opts_chunk$set(message = FALSE, fig.align = 'center', fig.width = 8)

theme_set(theme_pubr())
```

```{r}
#### import & set up data ####

# from https://github.com/eric-pedersen/groundfish-data-analysis
# note: these are mean biomasses over the whole sampled area with an SE
# not actually raw abundances. But for this purpose, should be ok
load("~/Documents/GitHub/groundfish-data-analysis/data/year_geom_means.Rdata")

time = rownames(Year_Geom_Means_all) %>% as.numeric()

# get number of populations & length of time series
npops <- ncol(Year_Geom_Means_all)
tsl <- nrow(Year_Geom_Means_all)
```

## Building the model with mgcv

```{r, results='hide'}
# scale all columns
Year_Geom_Means_all = apply(Year_Geom_Means_all, 2, scale) %>% as.data.frame()
Year_Geom_Means_all$time = as.integer(time) 
df = Year_Geom_Means_all %>% 
  pivot_longer(cols = -time, 
               names_to = "species", "values_to" = "N")
df$species <- factor(df$species)

# build a GI model: global smoother plus group-level smoothers with differing wiggliness

m <- gam(N ~ 
               # global smooth
               s(time, bs = c("tp"), m = 2, k = 30) + 
               # smooth per species
               s(time, by = species, bs = c("tp"), m = 1, k = 30) +
               # random intercept
               s(species, bs = c("re")), 
             data = df, 
             method="REML", 
             family="gaussian")

# save model
saveRDS(m, here("models/hgam.RDS"))
```

### Model performance

```{r}
gam.check(m)
k.check(m)

# draw(m)

summary(m)
```

## Predicting the full model

```{r}
pred = predict_gam(m)

ggplot(data = pred, aes(x = time, 
                        y = fit,
                        group = species)) +
 geom_smooth_ci(lwd = .3)
```

### Predict global smooth

```{r}
pred = predict_gam(m, exclude_terms = )

ggplot(data = pred, aes(x = time, 
                        y = fit,
                        group = species)) +
 geom_smooth_ci(lwd = .3)
```

## Extracting model coefficients: alpha

## Extract covariance between populations

## Extract global population trend

Note: this is *just* the temporal trend, without the latent variation or the residual error.

### Average slope of the temporal trend in population size (LPI-pop)

### Variance in the temporal trend across populations (LPI-comm)

```{r}
# drawing posteriors for the predicted response values:

beta <- coef(m)   # vector of model parameters

Vb <- vcov(m)     # default is the bayesian covariance matrix
heatmap(Vb, Rowv = NA, Colv = NA)

# Draw the estimated smooth 
# plot(m, seWithMean = TRUE, unconditional = TRUE)

# Evaluate the smooth over a range of values for `year`
time_sm <- smooth_estimates(m, "s(time)")
# add a confidence interval
time_sm <- confint(m, "s(time)")
time_sm

ggplot(data = filter(time_sm, smooth == "s(time)"), 
       aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = .4) +
  geom_line(aes(y = est)) +
  labs(y = "N", x = "") +
  ggpubr::theme_pubr()

ggplot(data = time_sm, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = smooth), alpha = .1) +
  geom_line(aes(y = est, col = smooth)) +
  geom_ribbon(data = filter(time_sm, smooth == "s(time)"),
              aes(x = time, ymin = lower, ymax = upper), alpha = .4) +
  geom_line(data = filter(time_sm, smooth == "s(time)"),
            aes(x = time, y = est)) +
  labs(y = "N", x = "") +
  ggpubr::theme_pubr() +
  theme(legend.position = "none") 


# sample from the posterior distribution
m_post <- smooth_samples(m, 's(time)', n = 100, seed = 42) 

post_summary <- m_post %>% 
  group_by(.x1) %>% 
  summarise(mu = mean(value, na.rm = TRUE),
            lower = quantile(value, 0.025),
            upper = quantile(value, 0.975),
            variance = var(value)) %>% ungroup()

time_fit <- evaluate_smooth(m, 's(time)')

draw(time_fit) + 
  geom_line(data = m_post, 
            aes(x = .x1, y = value, group = draw),
            alpha = 0.3, colour = 'red') 
# why does the global smooth start out at a higher intercept?
```

