# Hierarchical linear model on fish population trends

Model we are attempting to build:

$$N_{i,t} = f(t, \alpha_i) + g(\lambda_i, \beta_i) + \varepsilon_i$$

Where:

- $N$ : population size estimate
- $i$ : population
- $t$ : time step

- $f(t, \alpha_i)$ : function estimating the temporal trend in population size
  - $\alpha$: estimated parameter describing each population's temporal slope
- $g(\lambda, \beta_i)$ : function estimating latent variation in population size
  - $\lambda$ : latent variable (to capture covariation between populations)
  - $\beta$ : estimated parameter describing each population's latent variation
- $\varepsilon$: residual variation

Because the model also estimates latent variation, we are accounting for the covariation between populations when estimating $N$. 

This model then contains key information to calculate and extend an index like the Living Planet Index, to show the average change in population size from a baseline value of 1 in the start year.

The first part of the model estimates the temporal trend of each population's size ($\alpha_i$), which contains information about population-level change _and_ community-level change:

$$\alpha_i \sim N(\mu_{\alpha}, \sigma^2_{\alpha}) $$

where:
- \mu_{\alpha} is the average population trend
- \sigma^2_{\alpha} is the variance in this average trend, giving us a measure of how variable the population trends are within the sample.

To calculate the index of change from a baseline of 1 (as in the LPI), the average population trend can then be plugged in to the LPI calculation as follows:

$$ I_t = {\mu_{\alpha}}*t + 1 $$

This trend can then be visualized and communicated with its known ($\sigma^2_{\alpha}$).

## Setting up

```{r, include=FALSE}
library(here)
library(dplyr)
library(HMSC)
library(ggplot2)
library(ggpubr)

knitr::opts_chunk$set(message = FALSE, fig.align = 'center', fig.width = 8)

theme_set(theme_pubr())
```

```{r}
#### import & set up data ####

# from https://github.com/eric-pedersen/groundfish-data-analysis
# note: these are mean biomasses over the whole sampled area with an SE
# not actually raw abundances. But for this purpose, should be ok
load("~/Documents/GitHub/groundfish-data-analysis/data/year_geom_means.Rdata")

time = rownames(Year_Geom_Means_all) %>% as.numeric()

# get number of populations & length of time series
npops <- ncol(Year_Geom_Means_all)
tsl <- nrow(Year_Geom_Means_all)
```

## Building the model with HMSC

```{r, results='hide'}
# format data
XData <- as.matrix(time)
YData <- as.matrix(Year_Geom_Means_all) %>% apply(2, scale, center = FALSE)
randomEff <- as.factor(1:tsl) # assign a location to pairings of pops

# prepare data
formDat <- as.HMSCdata(X = XData, Y = YData, Random = randomEff, interceptX = TRUE)

# run model
m <- hmsc(formDat,
                family = "gaussian",
                niter = 15000,
                nburn = 5000,
                thin = 5,
                verbose = FALSE)
# save model
saveRDS(m, here("models/hmsc.RDS"))
```

### Model performance

```{r}
# extract the parameters and format them into an `mcmc` object.
mcmcMeansParamX <- as.mcmc(m, parameters = "meansParamX")

# check for parameter convergence (visually)
traceplot(mcmcMeansParamX)
```

```{r}
# Explanatory power 

# (R2 <- Rsquared(m, averageSp = FALSE))
(R2comm <- Rsquared(m, averageSp = TRUE))
```

## Predicting the full model

```{r}
# Predict and plot model over time 

# prepare and format data
formPredDat = as.HMSCdata(X = XData, Random = randomEff)

# predict population size over time from the model
pred <- predict(m,
                newdata = formPredDat,
                type = "response")
```

```{r}
# Average N across populations
pred_mean = apply(pred, 1, mean)
pred_sd = apply(pred, 1, sd)
```

```{r, echo=F, warning=FALSE}
plot(pred_mean ~ XData, 
     col = "white", xlab = "years", ylab = "N",
     ylim = c(min(pred), max(pred)))
for(i in 1:ncol(pred)) {lines(pred[,i] ~ XData, lwd = .5)}
lines(pred_mean ~ XData, lwd = 2, col = "red")
lines(pred_mean + 1.96*pred_sd ~ XData, lwd = 2, lty= 2, col = "red")
lines(pred_mean - 1.96*pred_sd ~ XData, lwd = 2, lty= 2, col = "red")
```


## Extract model coefficients: alpha

```{r}
# extract the parameters and format them into an `mcmc` object.
coefs = coef(m)
```

```{r}
temp = as.data.frame(coefs$paramX)
temp$pop = rownames(temp)
temp$pop <- factor(temp$pop, levels = temp$pop[order(temp$x1)])
ggplot(data = temp) +
  geom_point(aes(y = pop, x = x1)) +
  labs(y = "", x = "alpha i") +
  geom_vline(xintercept = 0, lwd = .3) +
  coord_cartesian(xlim = c(-1, 1))
```

```{r}
ggdensity(temp, x = "x1", add = "mean", rug = TRUE) +
  coord_cartesian(xlim = c(-1, 1)) +
  labs(title = "Distribution of alpha",
       subtitle = paste0("Mean = ", 
                         signif(mean(temp$x1), digits = 3),
                        ", Variance = ", 
                        signif(var(temp$x1), digits = 3)),
       y = "Density", x = "Alpha")
```

## Extract covariance between populations

```{r, fig.width=12, fig.height = 12}
# extract all estimated species-to-species associations matrix
assoMat <- corRandomEff(m)

# Average
siteMean <- apply(assoMat[, , , 1], 1:2, mean)

# Plot as heatmap
corrplot::corrplot(siteMean, type = "lower",
                   method = "color", 
                   tl.cex = .4, tl.col = "black")
```

## Extract global population trend

Note: this is *just* the temporal trend, without the latent variation or the residual error.

### Average slope of the temporal trend in population size (LPI-pop)

```{r}
(lpipop = mean(coefs$paramX[,2]))
```

### Variance in the temporal trend across populations (LPI-comm)

```{r}
(lpicomm = var(coefs$paramX[,2]))
```

So after accounting for the covariation between populations, an average population in this ensemble declined in size by 7.8% every year but the temporal trends vary a lot between populations ($\sigma^2$ = 0.11, so SD = 0.34).