# Hierarchical linear model on fish population trends

Model we are attempting to build:

$$N = f(t, \alpha_i) + g(\lambda, \beta_i) + \epsilon$$

Where:

- $N$ : population size estimate
- $t$ : time step
- $i$ : population
- $f(t, \alpha_i)$ : function estimating the temporal trend in population size
  - $\alpha$: each population's estimated temporal slope
- $g(\lambda, \beta_i)$ : function estimating latent variation in population size
  - $\lambda$ : latent variable (to capture covariation between populations)
  - $\beta$ : estimated parameter describing each popultion's latent variation
- $\epsilon$: residual variation

This model then contains key information to calculate and extend an index like the Living Planet Index, to show the average change in population size from a baseline value of 1 in the start year.

To calculate the LPI, the estimated global change in population size $N$ can be plugged into the calculation to get the overall growth rate:

$$dt = log_{10} (\frac{N_t}{N_{t-1}}) $$

Which can then be used to calculate the LPI as:

$$ I_t = I_{t-1}*10^{dt} $$

Because the model also estimates latent variation, we are accounting for the covariation between populations when estimating the global trend. 


## Setting up

```{r, include=FALSE}
library(here)
library(dplyr)
library(HMSC)
library(ggplot2)
```

```{r}
#### import & set up data ####

# from https://github.com/eric-pedersen/groundfish-data-analysis
# note: these are mean biomasses over the whole sampled area with an SE
# not actually raw abundances. But for this purpose, should be ok
load("~/Documents/GitHub/groundfish-data-analysis/data/year_geom_means.Rdata")

time = rownames(Year_Geom_Means_all) %>% as.numeric()

# get number of populations & length of time series
npops <- ncol(Year_Geom_Means_all)
tsl <- nrow(Year_Geom_Means_all)
```

## Building the model with HMSC

```{r, results='hide'}
# format data
XData <- as.matrix(time)
YData <- as.matrix(Year_Geom_Means_all)
randomEff <- as.factor(1:tsl) # assign a location to pairings of pops

# prepare data
formDat <- as.HMSCdata(X = XData, Y = YData, Random = randomEff)

# run model
m <- hmsc(formDat,
                family = "gaussian",
                niter = 15000,
                nburn = 5000,
                thin = 5,
                verbose = FALSE)
# save model
saveRDS(m, here("models/hmsc.RDS"))
```

### Model performance

```{r}
# extract the parameters and format them into an `mcmc` object.
mcmcMeansParamX <- as.mcmc(m, parameters = "meansParamX")

# check for parameter convergence (visually)
traceplot(mcmcMeansParamX)
```

```{r}
# model explanatory power
(R2 <- Rsquared(m, averageSp = FALSE))
(R2comm <- Rsquared(m, averageSp = TRUE))
```

## Population trends ($f(t, \alpha_i)$)

```{r, echo=FALSE}
# average model per species across iterations
avg <- as.data.frame(apply(m$results$estimation$paramX, 1:2, mean))

# get 95% confidence intervals
ci_lo <- as.data.frame(apply(m$results$estimation$paramX, 1:2, quantile, probs = 0.025))
ci_hi <- as.data.frame(apply(m$results$estimation$paramX, 1:2, quantile, probs = 0.975))
```

```{r, echo=FALSE}
# plot all populations' models
ggplot(data = as.data.frame(XData), aes(x = V1)) +
  geom_abline(data = avg, 
              aes(slope = x1, intercept = Intercept),
              lwd = 0.5) +
  # geom_abline(data = ci_lo,
  #             aes(slope = x1, intercept = Intercept),
  #             lty = 2, col = "blue") +
  # geom_abline(data = ci_hi,
  #             aes(slope = x1, intercept = Intercept),
  #             lty = 2, col = "red") +
  ylim(c(-1,2)) +
  theme_classic()
```


### Covariance between populations

```{r}
# extract all estimated species-to-species associations matrix
assoMat <- corRandomEff(m)

# Average
siteMean <- apply(assoMat[, , , 1], 1:2, mean)

# build matrix of colours (red : neg, blue: pos)
siteDrawCol <- matrix(NA, nrow = nrow(siteMean),
                      ncol = ncol(siteMean))
siteDrawCol[which(siteMean < 0, arr.ind=TRUE)]<-"red"
siteDrawCol[which(siteMean > 0, arr.ind=TRUE)]<-"blue"

# plot as a heatmap
heatmap(siteMean, Rowv = NA, Colv = NA, col = siteDrawCol)
```

## Global population trend

```{r}
# Predict and plot model over time 

# prepare and format data
formPredDat = as.HMSCdata(X = XData, Random = randomEff)

# predict values
pred <- predict(m,
                newdata = formPredDat,
                type = "response")
```

```{r}
# Get mean trend across populations

# take the mean N per year
pred_mean = apply(pred, 1, mean)
```

```{r, echo=F}
plot(pred_mean ~ XData, 
     col = "white", xlab = "years", ylab = "N",
     ylim = c(min(pred), max(pred)))
for(i in 1:ncol(pred)) {lines(pred[,i] ~ XData, lwd = .5)}
lines(pred_mean ~ XData, lwd = 2)
```

## Rates of population change

```{r, echo=F}
#### calculate rate of change ####

# calculate log-ratio population change
calc_dt <- function(N){
  dt = c(1) # initial value
  for(i in 2:length(N)){
    dt[i] = log10(N[i]/N[i-1])
  }
  return(dt)
}

dt = apply(pred, 2, calc_dt)

dt_bar = apply(dt, 1, mean, na.rm = TRUE)
dt_ci = apply(dt, 1, quantile, probs = c(0.025, 0.975), na.rm = TRUE)
dt_var = apply(dt, 1, var, na.rm = TRUE)

calc_lpi <- function(rate){
  lpi = 1
  for(t in 2:tsl){
    lpi[t] = lpi[t-1]*10^rate[t]
  }
  return(lpi)
}

lpi = calc_lpi(dt_bar)
lpi_pops = apply(dt, 2, calc_lpi)

# plot lpi and all population lpis
matplot(lpi_pops, x = time, 
        type = "l", lwd = 0.2, lty = 1, 
        ylim = c(0, 5))
lines(x = time, y = lpi, lwd = 3)

# plot lpi with CI from the variance
plot(x = time, y = lpi, type = "l", ylim = c(0,2))
lines(x = time, y = lpi - 1.96*sqrt(dt_var), lty = 2)
lines(x = time, y = lpi + 1.96*sqrt(dt_var), lty = 2)

# plot the variance
plot(x = time, y = dt_var, type = "l",
     ylab = "Variance of growth rates")
```

