---
theme: litera
---

# Hierarchical GAM on population trends in Bayesian framework

This is the model we are attempting to build, to apply to an ensemble of population time series:

$$N_{i,t} = f(t, \alpha_i) + g(\lambda_i, \beta_i) + \varepsilon_i$$

Where:

- $N$ : population abundance estimate
- $i$ : population
- $t$ : time step

- $f(t, \alpha_i)$ : function estimating the temporal trend in population size
  - $\alpha$: estimated parameter describing each population's temporal slope
- $g(\lambda, \beta_i)$ : function estimating latent variation in population size
  - $\lambda$ : latent variable (to capture covariation between populations)
  - $\beta$ : estimated parameter describing each population's latent variation
- $\varepsilon$: residual variation

Because the model also estimates latent variation, we are accounting for the covariation between populations when estimating $N$. 

This model then contains key information to calculate and extend an index like the Living Planet Index, to show the average change in population size from a baseline value of 1 in the start year.

The first part of the model estimates the temporal trend of each population's size ($\alpha_i$), which contains information about population-level change _and_ community-level change:

$$\alpha_i \sim N(\mu_{\alpha}, \sigma^2_{\alpha}) $$

where:
- \mu_{\alpha} is the average population trend
- \sigma^2_{\alpha} is the variance in this average trend, giving us a measure of how variable the population trends are within the sample.

To calculate the index of change from a baseline of 1 (as in the LPI), the average population trend can then be plugged in to the LPI calculation as follows:

$$ I_t = {\mu_{\alpha}}*t + 1 $$

This trend can then be visualized and communicated with its known ($\sigma^2_{\alpha}$).


```{r, include = FALSE}
library(here)
library(dplyr)
library(tidyr)
library(mgcv)
library(tidymv)
library(gratia)
library(ggplot2)
library(ggpubr)
#devtools::install_github("nicholasjclark/mvgam")
#remotes::install_github("stan-dev/cmdstanr")
library(cmdstanr)
library(mvgam)

knitr::opts_chunk$set(message = FALSE, fig.align = 'center', fig.width = 8, warning = FALSE)

theme_set(theme_pubr())
```

# Simulate discrete time series of population abundances

```{r}
# Import data
set.seed(2)
dat = readRDS("data/exampledata-CAMP2015.rds")

dat_train = filter(dat, time > 2003 & time < 2008)
dat_test = filter(dat, time >= 2008)

ggplot(dat_train) +
  geom_line(aes(x = time, y = y, group = ID, col = Binomial), lwd = .5)
ggplot(dat_train) +
  geom_line(aes(x = time, y = log10(y+1), group = ID, col = Binomial), lwd = .5)
```

## Building the model 

### Preparing the prior

```{r, eval = F}
mvgam_prior <- mvgam(data = dat_train,
               formula = y ~ s(time, bs = 'tp', k = 3) + s(ID, bs = 're'),
               family = 'poisson',
               trend_model = 'AR1',
               chains = 3,
               use_stan = TRUE,
               prior_simulation = TRUE)

plot(mvgam_prior, type = 'smooths', realisations = TRUE)
```

Checking the default prior distributions for this model formulation:

```{r}
test_priors <- get_mvgam_priors(y ~ s(time, bs = 'tp', k = 20) + s(series, bs = 're'),
                                family = 'poisson',
                                data = dat_train,
                                trend_model = 'AR1',
                                use_stan = TRUE)
test_priors %>% kableExtra::kable() %>% kableExtra::kable_styling(latex_options = "striped")
```

Any of the above priors can be changed by modifying the `prior` column and supplying the resulting `data.frame` to the `priors` argument in `mvgam()`.

### Training the model on data

Using the default priors:

```{r, results='hide', cache = TRUE}
mod1 <- mvgam(data = dat_train,
                newdata = dat_test,
                formula = y ~ s(time, bs = "tp", k = 20) + s(series, bs = 're'),
                use_lv = TRUE,
                n_lv = 1,
                family = 'poisson',
                trend_model = 'AR1',
                use_stan = TRUE,
                chains = 3,
                burnin = 100,
                samples = 2000)

# to view the Stan model file:
# code(mod1)
```

```{r}
plot(mod1)
```

### Trace plot

```{r}
rstan::stan_trace(mod1$model_output, 'rho')
```

For the latent trend:

```{r}
rstan::stan_trace(mod1$model_output, 'ar1')
```


### Posterior checks

Is the model able to simulate data for the training period that looks realistic and unbiased?

First, looking at histograms for the posterior retrodictions to compare to the histogram of observations (i.e. how well does the model predict the observed data?)

```{r}
for(i in 1:10){
  ppc(mod1, series = i, type = 'hist')
}
```

Then, comparing the simulated empirical Cumulative Distribution Functions (CDF) for posterior retrodictions (`yhat`) to the CDF of the observations (`y`):

```{r}
for(i in 1:10){
  ppc(mod1, series = i, type = 'cdf')
}
```

Then looking for biases in predictions, by examining a Probability Integral Transform (PIT) histogram: if our predictions are not biased one way or another (i.e. not consistently under- or over-predicting), this histogram should look roughly uniform:

```{r}
for(i in 1:10){
  ppc(mod1, series = i, type = 'pit')
}
```


### Plot the time smooth

```{r}
for(i in 1:10){
  plot_mvgam_smooth(object = mod1, series = i, smooth = 'time')
}
```


### Look at the correlations between "species"

Thanks to the latent variable, we can pull out the correlations between species' abundance trends:

```{r}
sp_correlations = lv_correlations(mod1)

# Plot as heatmap
corrplot::corrplot(sp_correlations$mean_correlations, 
                   type = "lower",
                   method = "color", 
                   tl.cex = .4, tl.col = "black")
```

```{r, include=FALSE, eval = TRUE}
# pdf("~/Documents/GitHub/hierarchical-lpi/figures/mvgam_mod1_uncertainty.pdf", onefile = TRUE)
# for(i in 1:30){
#   plot_mvgam_uncertainty(mod1, series = i, newdata = dat_test, legend_position = "none")
# }
# dev.off()
# 
# pdf("~/Documents/GitHub/hierarchical-lpi/figures/mvgam_mod1_trends.pdf", onefile = TRUE)
# for(i in 1:30){
#   plot_mvgam_trend(mod1, series = i, newdata = dat_test, legend_position = "none")
# }
# dev.off()
```

