# Hierarchical GAM on fish population trends in Bayesian framework

Model we are attempting to build:

$$N_{i,t} = f(t, \alpha_i) + g(\lambda_i, \beta_i) + \varepsilon_i$$

Where:

- $N$ : population size estimate
- $i$ : population
- $t$ : time step

- $f(t, \alpha_i)$ : function estimating the temporal trend in population size
  - $\alpha$: estimated parameter describing each population's temporal slope
- $g(\lambda, \beta_i)$ : function estimating latent variation in population size
  - $\lambda$ : latent variable (to capture covariation between populations)
  - $\beta$ : estimated parameter describing each population's latent variation
- $\varepsilon$: residual variation

Because the model also estimates latent variation, we are accounting for the covariation between populations when estimating $N$. 

This model then contains key information to calculate and extend an index like the Living Planet Index, to show the average change in population size from a baseline value of 1 in the start year.

The first part of the model estimates the temporal trend of each population's size ($\alpha_i$), which contains information about population-level change _and_ community-level change:

$$\alpha_i \sim N(\mu_{\alpha}, \sigma^2_{\alpha}) $$

where:
- \mu_{\alpha} is the average population trend
- \sigma^2_{\alpha} is the variance in this average trend, giving us a measure of how variable the population trends are within the sample.

To calculate the index of change from a baseline of 1 (as in the LPI), the average population trend can then be plugged in to the LPI calculation as follows:

$$ I_t = {\mu_{\alpha}}*t + 1 $$

This trend can then be visualized and communicated with its known ($\sigma^2_{\alpha}$).


```{r, include = FALSE}
# Script to build hierarchical GAM on fish population trends

library(here)
library(dplyr)
library(tidyr)
library(mgcv)
library(tidymv)
library(gratia)
library(ggplot2)
library(ggpubr)
#devtools::install_github("nicholasjclark/mvgam")
#remotes::install_github("stan-dev/cmdstanr")
library(cmdstanr)
library(mvgam)

knitr::opts_chunk$set(message = FALSE, fig.align = 'center', fig.width = 8)

theme_set(theme_pubr())
```

```{r}
#### import & set up data ####

# from https://github.com/eric-pedersen/groundfish-data-analysis
# note: these are mean biomasses over the whole sampled area with an SE
# not actually raw abundances. But for this purpose, should be ok
load("~/Documents/GitHub/groundfish-data-analysis/data/year_geom_means.Rdata")

year = rownames(Year_Geom_Means_all) %>% as.numeric()

# scale all columns
Year_Geom_Means_all = apply(Year_Geom_Means_all, 2, scale) %>% as.data.frame()
Year_Geom_Means_all$year = year 
df = Year_Geom_Means_all %>% 
  pivot_longer(cols = -year, 
               names_to = "species", 
               values_to = "y")
df$species <- factor(df$species)

# get number of populations & length of time series
npops <- ncol(Year_Geom_Means_all)
tsl <- nrow(Year_Geom_Means_all)

set.seed(2)
# dat <- sim_mvgam(T = 30, 
#                  n_series = 20, 
#                  n_lv = 1,
#                  family = 'poisson',
#                  mu_obs = runif(6, 4, 8),
#                  trend_rel = 0.6, 
#                  train_prop = 0.85)
dat_train = filter(df, year %in% c(1981:2000))
dat_test = filter(df, year %in% c(2001:2013))

ggplot(dat_train) +
  geom_line(aes(x = time, y = y, group = series), lwd = .1)
```

## Building the model 

```{r, results='hide'}
mod1 <- mvgam(data = dat$data_train,
                newdata = dat$data_test,
                formula = y ~ s(series, bs = 're'),
                knots = list(season = c(0.5, 12.5)),
                use_lv = TRUE,
                n_lv = 2,
                family = 'poisson',
                trend_model = 'RW',
                use_stan = TRUE,
                chains = 4,
                burnin = 300,
                samples = 400)
```

```{r}
plot(mod1)
```

```{r, results='hide'}
mvgam_prior <- mvgam(data = df,
               formula = y ~ s(time, bs = 'tp', k = 15),
               family = 'poisson',
               trend_model = 'AR1',
               chains = 2,
               use_stan = TRUE)

plot(lynx_mvgam_prior, type = 'smooths', realisations = TRUE)
# build a GI model: global smoother plus group-level smoothers with differing wiggliness

m <- gam(N ~ # global smooth
           s(time, bs = c("tp"), m = 2, k = 15) +
           # smooth per species
           s(time, by = species, bs = c("tp"), m = 1, k = 15) +
           # random intercept
           s(species, bs = c("re")),
         data = df,
         method="REML",
         family="gaussian")

# save model
saveRDS(m, here("models/hgam_modelGI.RDS"))
```

### Model performance

```{r}
m <- readRDS("models/hgam_modelGI.RDS")

gam.check(m)
# draw(m)
summary(m)
```

## Predicting the full model

### Predict global smooth (overall temporal trend)

```{r}
# plot the global smooth s(time)
plot(m, 
     residuals = TRUE, shade = TRUE, shade.col = "lightblue", 
     select = 1, 
     seWithMean = TRUE, 
     shift = unname(coef(m)[1]))
```

### Predict species smooths

```{r}
# Evaluate the smooth over a range of values for `year`
time_smest <- smooth_estimates(m, "s(time)", n = tsl)
# add a confidence interval
time_sm <- confint(m, "s(time)", n = tsl)

ggplot(data = time_sm, aes(x = time)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = smooth), alpha = .1) +
  geom_line(aes(y = est, col = smooth)) +
  geom_ribbon(data = filter(time_sm, smooth == "s(time)"),
              aes(x = time, ymin = lower, ymax = upper), alpha = .4) +
  geom_line(data = filter(time_sm, smooth == "s(time)"),
            aes(x = time, y = est)) +
  labs(y = "s(time)", x = "") +
  ggpubr::theme_pubr() +
  theme(legend.position = "none") 
```

## Extracting model coefficients: alpha

```{r}
time_smest

plot(x = time_smest$time, y = time_smest$est, type = "l")
lines(x = time_smest$time, 
      y = time_smest$est + time_smest$se, lty = 2)
lines(x = time_smest$time, 
      y = time_smest$est - time_smest$se, lty = 2)
```

## Extract covariance between populations

```{r}
Vb <- vcov(m, freq = FALSE) # default is the bayesian cov matrix
# heatmap(Vb, Rowv = NA, Colv = NA)
# corrplot::corrplot(Vb,
#                    method = "color",
#                    type = "lower",
#                    tl.pos = "n")

# Vb_sponly <- Vb[which(grepl("species", rownames(Vb))),
#                 which(grepl("species", colnames(Vb)))]
# Vb_sponly_ls <- list()
# for(i in 1:30){
#   Vb_sponly_ls[[i]] <- Vb_sponly[
#     which(grepl(paste0(i), rownames(Vb_sponly))),
#     which(grepl(paste0(i), colnames(Vb_sponly)))]
# }
# 
# Vb_cor = cov2cor(Vb_sponly)
# 
# for(i in 1:29){
# corrplot::corrplot(Vb_sponly_ls[[i]], 
#                    method = "color", 
#                    type = "lower", 
#                    tl.pos = "n")
# }
```

## Extract global population trend

Note: this is *just* the temporal trend, without the latent variation or the residual error.

### Average slope of the temporal trend in population size (LPI-pop)

```{r}
lpipop = filter(time_sm, smooth == "s(time)")$est
```


### Variance in the temporal trend across populations (LPI-comm)

```{r}
lpicomm = filter(time_sm, smooth == "s(time)")$se
# not sure about this one...
```

```{r}
# drawing posteriors for the predicted response values:

beta <- coef(m)   # vector of model parameters

Vb <- vcov(m)     # default is the bayesian covariance matrix
heatmap(Vb, Rowv = NA, Colv = NA)


# sample from the posterior distribution
m_post <- smooth_samples(m, 's(time)', n = 100, seed = 42) 

post_summary <- m_post %>% 
  group_by(.x1) %>% 
  summarise(mu = mean(value, na.rm = TRUE),
            lower = quantile(value, 0.025),
            upper = quantile(value, 0.975),
            variance = var(value)) %>% ungroup()

time_fit <- evaluate_smooth(m, 's(time)', n = tsl)

draw(time_fit) + 
  geom_line(data = m_post, 
            aes(x = .x1, y = value, group = draw),
            alpha = 0.3, colour = 'red') 


preds = predict(m, se.fit = TRUE, type = "response")
pred <- bind_cols(preds) %>% bind_cols(df)

ilink = inv_link(m)
crit <- qnorm((1 - 0.89) / 2, lower.tail = FALSE) # or just `crit <- 2`
pred <- mutate(pred, Npred = ilink(fit),
               lwr = ilink(fit - (crit * se.fit)), # lower...
               upr = ilink(fit + (crit * se.fit))) # upper credible interval

ggplot(pred, aes(x = time, group = species)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
    geom_line(aes(y = Npred), lwd = .2) + 
  labs(y = "Predicted population size", x = NULL)



beta <- coef(m)   # vector of model parameters
Vb <- vcov(m)     # default is the Bayesian covariance matrix
Xp <- predict(m, type = 'lpmatrix')
set.seed(42)
beta_sim <- rmvn(n = 1000, beta, Vb) # simulate parameters
eta_p <- Xp %*% t(beta_sim)        # form linear predictor values
mu_p <- inv_link(m)(eta_p)    # apply inverse link function
mean(mu_p[1, ]) # mean of posterior for the first observation in the data
quantile(mu_p[1, ], probs = c(0.025, 0.975))


year1_row <- which(df$time == 1981)
hist(mu_p[year1_row,], main = "1990")
```

