To calculate the LPI, the estimated global change in population size $N$ can be plugged into the calculation to get the overall growth rate:

$$dt = log_{10} (\frac{N_t}{N_{t-1}}) $$

Which can then be used to calculate the LPI as:

$$ I_t = I_{t-1}*10^{\bar{dt}} $$

The confidence intervals for the index can be obtained directly from the residual error in the model estimating $N$, which propagates through the calculation of the log-ratio growth rate into the transformation into the index of change $I_t$. 


## Visualize the population temporal trends ($f(t, \alpha_i)$)

```{r, echo=FALSE}
# average model per species across iterations
avg <- as.data.frame(apply(m$results$estimation$paramX, 1:2, mean))

# get 95% confidence intervals
ci_lo <- as.data.frame(apply(m$results$estimation$paramX, 1:2, quantile, probs = 0.025))
ci_hi <- as.data.frame(apply(m$results$estimation$paramX, 1:2, quantile, probs = 0.975))
```

```{r, echo=FALSE}
# plot all populations' models
ggplot(data = as.data.frame(XData)) +
  geom_abline(data = avg, 
              aes(slope = x1, intercept = Intercept),
              linewidth = 0.2) +
  # geom_abline(data = ci_lo,
  #             aes(slope = x1, intercept = Intercept),
  #             lty = 2, col = "blue") +
  # geom_abline(data = ci_hi,
  #             aes(slope = x1, intercept = Intercept),
  #             lty = 2, col = "red") +
  ylim(c(-1,1))
```







--------

## Predicting the full model

```{r}
# Predict and plot model over time 

# prepare and format data
formPredDat = as.HMSCdata(X = XData, Random = randomEff)

# predict population size over time from the model
pred <- predict(m,
                newdata = formPredDat,
                type = "response")
```

```{r}
# Average N across populations
pred_mean = apply(pred, 1, mean)
pred_sd = apply(pred, 1, sd)
```

```{r, echo=F, warning=FALSE}
plot(pred_mean ~ XData, 
     col = "white", xlab = "years", ylab = "N",
     ylim = c(min(pred), max(pred)))
for(i in 1:ncol(pred)) {lines(pred[,i] ~ XData, lwd = .5)}
lines(pred_mean ~ XData, lwd = 2, col = "red")
lines(pred_mean + 1.96*pred_sd ~ XData, lwd = 2, lty= 2, col = "red")
lines(pred_mean - 1.96*pred_sd ~ XData, lwd = 2, lty= 2, col = "red")
```

## Rates of population change

```{r, echo=T, warning=F}
# calculate log-ratio population change
calc_dt <- function(N){
  dt = c(0) # initial value
  for(i in 2:length(N)){
    dt[i] = log10(N[i]/N[i-1])
  }
  return(dt)
}

dt_bar = calc_dt(pred_mean)
```

```{r}
plot(x = time, y = dt_bar, type = "l", ylim = c(-1, 1.2))
```



## Calculate an index of average change (à la LPI)

```{r, echo=T}
calc_lpi <- function(rate){
  lpi = 1
  for(t in 2:tsl){
    lpi[t] = lpi[t-1]*10^rate[t]
  }
  return(lpi)
}

lpi = calc_lpi(dt_bar)
lpi_pops = apply(dt, 2, calc_lpi)
lpi_pops[which(is.nan(lpi_pops))] <- 0

# plot lpi and all population lpis
matplot(lpi_pops, x = time, 
        type = "l", lwd = 0.2, lty = 1, 
        ylim = c(0, 5),
        main = "LPI with each populations' individual LPI")
lines(x = time, y = lpi, lwd = 3)
```

### Show uncertainty in the mean

```{r, echo=F}

# function to calculate geometric mean that will be used in bootstrapping
dtmean <- function(dt, i){
  d <- dt[i] # indexing happens here
  avg <- mean(d, na.rm = TRUE)
  return(avg)
}

# function for bootstrap confidence interval 
dt_boot = function(dt){
  # bootstrap resampling
  dt_r = boot::boot(dt, statistic = dtmean, R = 1000)
  # calculate 95% confidence intervals
  dt_ci = boot::boot.ci(dt_r, type = "basic", R = 1000)
  # wrangle into table for output
  dtboot_df = data.frame(mu = dt_ci$t0, 
                         cilo = dt_ci$basic[4], 
                         cihi = dt_ci$basic[5])
  return(dtboot_df)
}

dt_ci = apply(dt, 1, dt_boot) %>% bind_rows()
lpi_lower = calc_lpi(dt_ci$cilo)
lpi_upper = calc_lpi(dt_ci$cihi)

# plot lpi with CI from the variance
plot(x = time, y = lpi, type = "l", ylim = c(0,1.2),
     main = "LPI of all populations")
abline(h = 1, lty = 2)
lines(x = time, y = lpi_upper, lty = 3)
lines(x = time, y = lpi_lower, lty = 3)
```

